---
title: "NOTES - US Storm Data Analysis"
author: "Helen"
date: "21 February 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE)
```

```{r, cache = TRUE}
noaa <- read.csv("StormData.csv.bz2")
str(noaa)
```

I also load the packages I may need.

```{r, warning = FALSE}
library(ggplot2)
library(tidyr)
library(dplyr)
```


## Ideas / Playing with Data

I tried putting the dates and times into a non-factor format but it was a pain in the butt due to missing data so decided it wasn't required.

The information I have been given states that we want to find out which events (in the EVTYPE variable) are most harmful to human health. However, there are 935 different types of events. I shall try to group these together to make the analysis more manageable.

Firstly, I know there are several types of "heat" - excessive heat, excess heat, heat etc. These also are responsible for all the incidences of a large number of deaths.

I will remove these into their own dataset and then create a remaining dataset without these.

```{r, cache = TRUE}
noaa_heatl <- grepl("heat", noaa$EVTYPE, ignore.case = TRUE)
noaa_heat <- noaa[noaa_heatl, ]
noaa_remain <- noaa[!noaa_heatl, ]
rm(noaa_heatl)
plot(noaa_heat$FATALITIES)
```

There are way too many of these. Let's look at the ones where people died.

```{r, cache = TRUE}
noaa_10f <- noaa[noaa$FATALITIES > 10, ]
summary(noaa_10f$EVTYPE)
```

This shows that incidents resulting in a high number of deaths only come from a few sources and a few of these are very similar (e.g. flood / flash flood).

Looking at those incidents that have killed at least 1 person:

```{r}
summary(noaa$EVTYPE[noaa$FATALITIES > 0])
```

The information I have been given states that we want to find out which events (in the EVTYPE variable) are most harmful to human health. However, there are 935 different types of events. I shall try to group these together to make the analysis more manageable.

I start by grouping everything the same way (as "Other") and then gradually identifying the top causes and categorising them.

Firstly let's add the extra column and then see what levels of fatalities the dataset includes.

```{r, cache = TRUE}
noaa <- tbl_df(noaa)
noaa <- mutate(noaa, EVTYPE_1 = "Other")
head(noaa$EVTYPE_1)
plot(noaa$FATALITIES)
```

50 looks like a good place to start.

```{r, cache = TRUE}
noaa_50f <- noaa[noaa$FATALITIES > 50, ]
noaa_50f <- droplevels(noaa_50f)
noaa_50f$EVTYPE
```

So there are four levels - three heats and one tornado. I'll add these categories.

```{r}
noaa$EVTYPE_1[grepl("heat", noaa$EVTYPE, ignore.case = TRUE)] <- "Heat"
noaa$EVTYPE_1[grepl("tornado", noaa$EVTYPE, ignore.case = TRUE)] <- "Tornado"
sum(noaa$EVTYPE_1 == "Heat")
```

So let's now plot the fatalities split by the EVTYPE.
